---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MIMIC3db

<!-- badges: start -->
<!-- badges: end -->

This work references data on from the Medical Information Mart for Intensive Care MIMIC-III database v1.4. 
MIMIC-III is a large, freely-available database comprising de-identified health-related data from patients who
were admitted to the critical care units of the Beth Israel Deaconess Medical Center from 2001-2019.  Detailed
information can be obtained on the [MIMIC-III website](https://mimic.mit.edu/docs/iii/).

MIMIC3db provides a tightly bound set of routines to build a SQLite version of the 
[MIMIC-III v1.4](https://physionet.org/content/mimiciii/1.4/) database on a local network.  The package has routines 
that then assist the user with loading the MIMIC-III data into the local database.  The package also presents base
routines for accessing data from the database.

## Installation

You can install MIMIC3db from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("hdshea/MIMIC3db")
```

## Basic Usage

Assuming that you have loaded the appropriate version of the MIMIC-III database into a database at 
`db/MIMIC-III.db`, then the following code will set up a connection `con` to that database.

```{r source_aux_functions, include=TRUE}
library(MIMIC3db)

base_dir <- here::here("")
db_file <- fs::path(base_dir, "db/MIMIC-III.db")
if(RSQLite::dbCanConnect(RSQLite::SQLite(), db_file)) {
    con <- RSQLite::dbConnect(RSQLite::SQLite(), db_file)
} else {
    stop(stringr::str_c("Database file: ", db_file, " not found.", sep=""))
}
```

## Building the database

Assuming you are in the base directory of an RStudio project that has a `db` directory defined in it and that you have loaded the SQL scripts from this GitHub directory, the following code run in the terminal from the base directory 
will create the base tables for the MIMIC-III v1.4 database and tables.

```{bash build_database, echo = TRUE, eval = FALSE}
cd db
sqlite3 MIMIC-III.db < mimic3_create_scipt.sql
cd ..
```


This is a basic example showing the use of some base function with an in memory example database:

```{r example}
library(MIMIC3db)
con <- RSQLite::dbConnect(RSQLite::SQLite(), ":memory:")
p <- tibble::tibble(SUBJECT_ID = c(10006, 10011))
a <- tibble::tibble(SUBJECT_ID = c(10006, 10011))
RSQLite::dbWriteTable(con, "PATIENTS", p)
RSQLite::dbWriteTable(con, "ADMISSIONS", a)

patients <- db_get_from_table(con, "PATIENTS")
patients

admissions <- db_get_from_table(con, "ADMISSIONS", where = "WHERE SUBJECT_ID = 10006")
admissions

RSQLite::dbDisconnect(con)
```


