---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MIMIC3db

<!-- badges: start -->
[![R-CMD-check](https://github.com/hdshea/MIMIC3db/workflows/R-CMD-check/badge.svg)](https://github.com/hdshea/MIMIC3db/actions)
<!-- badges: end -->

This work references data from the Medical Information Mart for Intensive Care MIMIC-III database v1.4. 
MIMIC-III is a large, freely-available database comprising de-identified health-related data from patients who
were admitted to the critical care units of the Beth Israel Deaconess Medical Center from 2001-2019.  Detailed
information can be obtained on the [MIMIC-III website](https://mimic.mit.edu/docs/iii/).

MIMIC3db provides a tightly bound set of routines to reference a [SQLite](https://www.sqlite.org/index.html) 
version of the [MIMIC-III v1.4](https://physionet.org/content/mimiciii/1.4/) database on a local network using 
base access routines from the [RSQLite](https://github.com/r-dbi/RSQLite) R package. 

## Installation

You can install MIMIC3db from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("hdshea/MIMIC3db")
```

## Building the database

Assumptions:

1. you are in the base directory of an RStudio project that has a `db` directory defined in it,
2. you have loaded the SQL scripts from 
[this GitHub directory](https://github.com/hdshea/MIMIC3db/tree/main/inst/sql) into the `db` directory, and
3. you have loaded the MIMIC-III v1.4 raw data files into the `db` directory retaining their default names

NOTE: you can put the database and the SQL scripts in any directory that you like; this example just uses a 
directory named `db` in the base project directory as an example.

These assumptions having been met, the following code run in the terminal from the base directory 
will:

1. create the database and base tables for MIMIC-III v1.4
2. load the data into the tables, and 
3. clean up some data columns and create base indexes for the tables.

NOTE:  the zip file containing the MIMIC-III v1.4 raw data files is 6.2 GB and the built database on my 
MacBook Pro M1 local hard drive is 85.6 GB.

```{bash build_database, echo=TRUE, eval=FALSE}
cd db
sqlite3 MIMIC-III.db < mimic3_create_script.sql
sqlite3 MIMIC-III.db < mimic3_load_small_tables.sql
sqlite3 MIMIC-III.db < mimic3_load_CHARTEVENTS.sql
sqlite3 MIMIC-III.db < mimic3_load_DATETIMEEVENTS.sql
sqlite3 MIMIC-III.db < mimic3_load_INPUTEVENTS_CV.sql
sqlite3 MIMIC-III.db < mimic3_load_INPUTEVENTS_MV.sql
sqlite3 MIMIC-III.db < mimic3_load_LABEVENTS.sql
sqlite3 MIMIC-III.db < mimic3_load_NOTEEVENTS.sql
sqlite3 MIMIC-III.db < mimic3_load_OUTPUTEVENTS.sql
sqlite3 MIMIC-III.db < mimic3_load_PRESCRIPTIONS.sql
sqlite3 MIMIC-III.db < fix_data_fields.sql
sqlite3 MIMIC-III.db < mimic3_add_indexes.sql
cd ..
```

This code then shows the tables in the new database.

```{r pre_connect_internal, include=FALSE, eval=TRUE}
db_file <- system.file("extdata", "MIMIC-III.db", package = "MIMIC3db")
if(RSQLite::dbCanConnect(RSQLite::SQLite(), db_file)) {
    con <- RSQLite::dbConnect(RSQLite::SQLite(), db_file)
} else {
    stop(stringr::str_c("Database file: ", db_file, " not found.", sep=""))
}
```

```{r pre_connect, include=TRUE, eval=FALSE}
base_dir <- here::here("")
db_file <- fs::path(base_dir, "db/MIMIC-III.db")
if(RSQLite::dbCanConnect(RSQLite::SQLite(), db_file)) {
    con <- RSQLite::dbConnect(RSQLite::SQLite(), db_file)
} else {
    stop(stringr::str_c("Database file: ", db_file, " not found.", sep=""))
}
```

```{r list_tables, include=TRUE, eval=TRUE}
RSQLite::dbListTables(con)
```

```{r pre_disconnect, include=TRUE, eval=TRUE}
RSQLite::dbDisconnect(con)
```

## Basic Usage

Assuming that you have loaded the appropriate version of the MIMIC-III database into a database at 
`db/MIMIC-III.db`, then the following code will set up a connection `con` to that database.

```{r connect_internal, include=FALSE, eval=TRUE}
library(MIMIC3db)

db_file <- system.file("extdata", "MIMIC-III.db", package = "MIMIC3db")
if(RSQLite::dbCanConnect(RSQLite::SQLite(), db_file)) {
    con <- RSQLite::dbConnect(RSQLite::SQLite(), db_file)
} else {
    stop(stringr::str_c("Database file: ", db_file, " not found.", sep=""))
}
```

```{r connect, include=TRUE, eval=FALSE}
library(MIMIC3db)

base_dir <- here::here("")
db_file <- fs::path(base_dir, "db/MIMIC-III.db")
if(RSQLite::dbCanConnect(RSQLite::SQLite(), db_file)) {
    con <- RSQLite::dbConnect(RSQLite::SQLite(), db_file)
} else {
    stop(stringr::str_c("Database file: ", db_file, " not found.", sep=""))
}
```

This is a basic example showing the use of the two base functions.

```{r base_functions, include=TRUE}
patients <- db_select_data(con, "SELECT * FROM PATIENTS")
patients

admissions <- db_get_from_table(con, "ADMISSIONS", where = "WHERE SUBJECT_ID = 10006")
admissions
```

This is a basic example showing the use of a few of the tables specific data access functions.

```{r table_functions, include=TRUE, eval=TRUE}
patients <- db_get_patients(con)
patients

admissions <- db_get_admissions(con, where = "WHERE SUBJECT_ID = 10006")
admissions
```

(NOTE:  The example database I use for testing has only empty tables.  That is why there are no results returned in the examples calls above.)

This is how you disconnect from the database when finished.

```{r disconnect, include=TRUE, eval=TRUE}
RSQLite::dbDisconnect(con)
```

## Citations for MIMIC-III publications:

Johnson, A., Pollard, T., & Mark, R. (2016). MIMIC-III Clinical Database (version 1.4). PhysioNet.
[https://doi.org/10.13026/C2XW26](https://doi.org/10.13026/C2XW26).

**Original publication**:
[Johnson, A. E. W., Pollard, T. J., Shen, L., Lehman, L. H., Feng, M., Ghassemi, M., Moody, B., Szolovits, P., 
Celi, L. A., & Mark, R. G. (2016). MIMIC-III, a freely accessible critical care database. Scientific Data, 3,
160035](https://www.nature.com/articles/sdata201635).

**PhysioNet**:
Goldberger, A., Amaral, L., Glass, L., Hausdorff, J., Ivanov, P. C., Mark, R., ... & Stanley, H. E. (2000).
PhysioBank, PhysioToolkit, and PhysioNet: Components of a new research resource for complex physiologic signals.
Circulation [Online]. 101 (23), pp. e215–e220.

